---
title: "Course Project 1"
author: "Akshay Avinash"
date: "13 September 2018"
output: html_document
---

### Loading and preprocessing the data  

Description of this data from course is as following :  
  
*This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.*  
  
There are three variables in the dataset:
  
1. **steps** : Number of steps taking in a 5-minute interval (missing values are coded as NA )
2. **date** : The date on which the measurement was taken in YYYY-MM-DD format
3. **interval** : Identifier for the 5-minute interval in which measurement was taken
  
Unzip and load the data into environment  
  
```{r loading}
unzip("activity.zip")
data <- read.csv("activity.csv")
```
  
-----  
### What is mean total number of steps taken per day?  
  
Create a histogram for the frequency of umber of steps taken  
Calculate the mean and median and show them in the graph  
  
```{r mean_of_steps}
totalSteps <- aggregate(data$steps,by=list(data$date),na.rm=T,FUN=sum)

mean_value <- round(mean(totalSteps$x),1)
median_value <- round(median(totalSteps$x),1)

hist(totalSteps$x,
     main = "Total Number of Steps Taken Per Day",
     breaks = 20,
     xlab = "Total Number of Steps")

abline(v=mean_value,lwd=2,col="blue")
abline(v=median_value,lwd=2,col="red")

legend('topright', lty = 1, lwd = 3, col = c("blue", "red"),
               cex = .8, 
               legend = c(paste('Mean: ', mean_value),
               paste('Median: ', median_value))
               )
```